---
title: "Race-Based Disparities in Federal Criminal Sentencing"
subtitle: "We'll think of one eventually"
author: "who knows"
date: "`r format(Sys.Date(),'%B %d, %Y')`"
output:
  github_document:
    toc: true
    number_sections: true
    # theme: yeti
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(fig.width=4, fig.height=4) 
```

---

This work is governed under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International license](https://creativecommons.org/licenses/by-nc-sa/4.0/).

Citation: Laurie Baker, Helen Grundman, Lily Khadjavi, Saeja Kim, Momin Malik, Ariana Mendible, Omayra Ortega, José Pabón, Chad M. Topaz, Thomas Wakefield. Institute for the Quantitative Study of Inclusion, Diversity, and Equity (QSIDE), with in-kind support from Institute for Computational and Experimental Research on Mathematics (ICERM), 2023.

---

# Acknowledgments

To be written.

# Background

```{r child = "Background.Rmd"}
```

```{r child = "sentencing-eda.Rmd"}
```

```{r child = "Exploration.Rmd"}
```

```{r child = "results.Rmd"}
```
