---
output: html_document
---

# Data exploration

Alright, let's start digging into the data. Whenever you are handed a new data set, you want to read in the data so you can view it in RStudio and perform an Exploratory Data Analysis (EDA). 

## Reading in the data

You can read in your data using `read_csv()` and creating a variable named `us_sent` that refers to the data set.

```{r read in data}
us_sent <- read_csv("data/cleaned_data_renamed.csv")
```

To view the data in a new tab within RStudio, you can use the `View()` function. Because this is an interactive RStudio function that doesn't work when we knit the document, we have set up our document so that the code can be viewed below but not actually run when we knit (specifically, we used the code chunk option `eval = FALSE`). 

```{r view data, eval = FALSE}
View(us_sent)
```

## Structure of the data

As you may remember, in a tidy data set, each row is an observation, each column is a variable, and each cell is a value. In the case of the federal criminal sentencing data, each row represents an individual who was sentenced in the federal district court system in the U.S. in one of the country's 94 districts.

When we start exploring a data set, we will want to think about some key questions using the four W's: who, what, when, and where. Specifically, you might ask: who is in our data set? What was the sentence? When was the individual was sentenced? Where did the sentence occur? 

Can you guess which variables pertain to each question? Let's inspect our variables, which are in the columns, and see if that helps. There are several functions we can use to do this:

* `names()` will provide the names of the columns in the data set
    ```{r check column names}
    names(us_sent)
    ```
    
* `summary()` will provide the names of the columns in the data set along with a brief numeric or qualitative summary of each variable

    ```{r check column summaries}
    summary(us_sent)
    ```

* `str()` will display (sometimes complex) details of the internal structure of the R object, including the table dimensions, variable names, variable types, and the first 10 observations for each variable within a data table

    ```{r check object structure}
    str(us_sent)
    ```
    
* `glimpse()` is a simplified version of `str()` that only displays the dimensions, variable names, variable types, and the first set of observations for each variable within a data table (how many observations depends on the width of your output)

    ```{r get glimpse of data}
    glimpse(us_sent)
    ```
    
> Note: You can use the built in R help documentation to learn more about each function by typing a question mark followed by the name of the function in the console, e.g. `?str()` 

## Data Types

What is halfway between 0 and 1? It is 1/2. What is halfway between horse and dog? There is no such thing! Thinking about each type of data is very important so that we don't code silly things like the mean of animal species.

There are two main types of data: **categorical data** and **numerical data**. Some examples of categorical data would be color, ethnicity, employment status, or states/countries. These have unique values, like California or Oregon. In other words, categorical variables take on discrete, finite values. 

On the other hand, numerical data is in number form. Some examples of numerical data are temperature, height, and salary. It makes perfect sense to be 165.8 cm tall or for the temperature to be 82.4 degrees outside.

There are some confusing data types that use numbers to *represent* categorical data, like zip code. You may live in the zip code 90201, which is a number, but you can't live in the zip code 90210.3. Only whole numbers, and specific ones at that, make sense here. We will learn more about using numbers to represent categorical data in this lesson.

Can you identify which data type each variable in the federal criminal sentencing data is?

## Explorations

A good practice to do with a new data set is to explore it through visualization. We will walk through a few graphs in R so you can see how to plot. We will examine each of these so we can see relationships and learn more about our data. Then, explore on your own by modifying this code!


## How to make a scatter plot with quantitative variables

In this first plot, we will look at numerical variables only using a scatter plot. Scatter plots help us to visualize and understand numerical data better. We will compare each defendant's age to their sentence length through visualization and observation, and we will use this scatter plot to identify any patterns that exist in our data set. Each dot in the scatter plot we produce with the `ggplot()` command from the **ggplot2** package represents a row in our `us_sent` data. 

To use `ggplot()`, we start by indicating what data set we want to plot, `us_sent`. Then, we sepcify what our variables of interest are--in this case, we are interested in `age` and `sentence_length`. `geom_point()` specifies that we are interested in using a scatter plot to represent our data. There are many fun and useful modifications that change the way a scatter plot looks. For example, you can change the color of data points using `color`, you can change the size of data points using `size`, and you can specify the transparency of data points using `alpha`. It's also good practice to come up with a useful title and good axis labels using the `labs` command.

```{r sentence length by age, fig.width = 6, fig.height = 6}
ggplot(us_sent, aes(x = age, 
                    y = sentence_length)) + 
  geom_point(color = 'coral2',
             size = 0.5,
             alpha = 0.4) +
  labs(title = "Sentence Length by Age",
       x = "Age (Years)",
       y = "Sentence Length (Months)")
```

What do you notice in this scatter plot? What do you wonder?

## What happens when you try to make a scatter plot with categorical variables?

We just created a scatter plot with two numerical variables. Now we will see what happens if one variable is numerical and the other is categorical. Run the code below that plots `criminal_history` (a categorical variable) against `age` (a numerical variable).

```{r criminal history by age, fig.width = 6, fig.height = 6}
ggplot(us_sent, aes(x = age, 
                    y = criminal_history)) + 
  geom_point(color = 'coral2', 
             alpha = 0.7) +
  labs(title = "Criminal History by Age",
       x = "Age (Years)",
       y = "Criminal History (1-6 levels)") # come back to this
```

What do you notice in this scatter plot? What do you wonder?

## What happens when you try to make a scatterplot with two catergoical variables?

Let's try one more scenario. What happens when you try to plot two categorical variables against each other? Run the code below to plot sex against species (both categorical variables).

```{r race by sex, fig.width = 6, fig.height = 6}
us_sent1 <- us_sent %>% 
  mutate(sex = as.factor(sex)) ## COME BACK TO THIS

ggplot(us_sent1, aes(x = race,
                    y = sex)) + 
  geom_point(color = 'coral2', 
             alpha = 0.7) +
  labs(title = "Race by Sex",
       x = "Race",
       y = "Sex (Male or Female)")
```

What do you notice in this scatter plot? What do you wonder?

As you may have noted, scatter plots of two categorical variables are not that useful for analysis and inference since they only display the way we've grouped our data and not any of the underlying patterns. To compare two categorical variables frequency tables or bar graphs are a better visualization to use. In the remainder of this lesson we will focus on comparisons where we have *at least one numerical variable*.


# Explanatory Data Analysis
<!-- - explain the different columns (linking to the background) -->
Now, let's return to exploring the variables of interest. Remember, we want to know *who* is in our data set, *what* the sentence was, *when* the individual was sentenced and *where* the sentence occurred. Did you identify how our variables might correspond with each of these questions? You might have classified the variables as follows: 

<!-- https://www.nature.com/articles/s41599-023-01879-5 -->

|Who | What| When | Where |
|:---|:---|:---|:---|
| `age` | `sentence_length` | `year` | `district` |
| `sex` | `mandatory_min` | | |
| `educ` | |||
| `race` | |||
| `criminal_history` | |||
| `guilty_plea` | |||

Let's start exploring the variables that fall under each of these key questions.

## Who

**Sex**

Remember that `sex` has been coded as a binary variable 0 and 1, where 0 is "Male" and 1 is "Female". We can confirm this using `distinct()`, a function that displays all unique values under a specified column. 

```{r}
us_sent %>%
  distinct(sex)
```

We can recode this variable using `mutate` and `case_when`, so that we simply see the words "Male" and "Female", rather than the coded numbers. Let's also use `ggplot()` to view the distribution of `sex` within our data set, this time using `geom_bar()` to display the data in a bar graph. Another difference from when we made a scatter plot above using `geom_point()` is that we are adding `fill` to make sure the bars in the bar graph are colored in. We can also add `scale_fill_viridis_d()` to automatically apply a certain color scheme to our plot.

```{r sex of sentenced individuals, fig.width = 5, fig.height = 5}
us_sent %>%
  mutate(sex = case_when(sex == 0 ~ "Male",
                         sex == 1 ~ "Female")) %>%
  ggplot() +
  geom_bar(aes(x = sex, fill = sex)) +
  labs(x = "Sex*",
       title = "Sex* of Sentenced Individuals",
       y = "Number of Individuals",
       fill = "Sex*") +
  scale_fill_viridis_d()
```

From this graph, we can see that there are far more males than females in our data set. 

**Race**

Let's continue exploring through turning toward the `race` variable. We will start our exploration of `race` by using `distinct`.

```{r}
us_sent %>%
  distinct(race)
```

We can once again use `ggplot()` to make a bar graph displaying the `race` variable. We can make use of `fct_infreq()` when choosing `race` as our x variable, which makes sure the bars in the plot are ordered by largest to smallest from left to right. 

```{r race of sentenced individuals, fig.width = 5, fig.height = 5}
us_sent %>%
  mutate(race = case_when(race == "other" ~ "ari",
                         TRUE ~ race)) %>%
  ggplot() +
  geom_bar(aes(x = fct_infreq(race), fill = race)) +
  labs(x = "Race*",
       title = "Race* of Sentenced Individuals",
       y = "Number of Individuals",
       fill = "Race*") +
  scale_fill_viridis_d()
```

**Race and Sex**

Now let's look at how `race` and `sex` relate to one another. First, let's look at the number of people in each combination using `count` again. 

```{r}
us_sent %>%
  mutate(sex = case_when(sex == 0 ~ "Male",
                         sex == 1 ~ "Female")) %>% 
  count(race, sex)
```

We can represent these two variables using a bar graph. Once again we will use `ggplot` to create this plot. This time, since we are creating a bar graph displaying two different variables, we can define `race` as the x variable and have the colors of the bar graph correspond to the `sex` variable. 

```{r, , fig.width = 5, fig.height = 5}
us_sent %>%
  mutate(race = case_when(race == "other" ~ "ari",
                         TRUE ~ race)) %>%
  mutate(sex = case_when(sex == 0 ~ "Male",
                         sex == 1 ~ "Female")) %>%
  ggplot() +
  geom_bar(aes(x = fct_infreq(race), fill = sex)) +
  labs(x = "Race*",
       title = "Race* and Sex* of Sentenced Individuals",
       y = "Number of Individuals",
       fill = "Sex*")  +
  scale_fill_viridis_d()
```

We can also create a proportional bar plot. All we need to add to our previous `ggplot()` is `position = "fill"` within `geom_bar()`. 

```{r, fig.width = 5, fig.height = 5}
us_sent %>%
  mutate(race = case_when(race == "other" ~ "ari",
                         TRUE ~ race)) %>%
  mutate(sex = case_when(sex == 0 ~ "Male",
                         sex == 1 ~ "Female")) %>%
  ggplot() +
  geom_bar(aes(x = fct_infreq(race), fill = sex), position = "fill") +
  labs(x = "Race*",
       title = "Race* and Sex* of Sentenced Individuals",
       y = "Number of Individuals",
       fill = "Sex*")  +
  scale_fill_viridis_d()
```

**Age**

We might want to explore what the age of different individuals is across districts. Let's explore the districts of Maine, Rhode Island, and Vermont. We can use the `filter` command to look at only these three districts of interest. 

```{r, eval = FALSE}
us_sent %>%
  filter(district %in% c("Maine", "Rhode Island", "Vermont")) %>%  
  ggplot() +
  geom_bar(aes(x = age, fill = district)) +
  labs(title = "Age in Maine, Rhode Island, and Vermont",
       x = "Age",
       y = "Count")
```


**To add**:
- what do we divide by?
- You may look at this plot and say that whites are sentenced at the federal district court more than black individuals. This is where it is important to look at the population in each district.


**Future Directions**

Bringing in the spatial district files

- https://www.openicpsr.org/openicpsr/project/100069/version/V1/view
- Downloading census data from the API.
- Joining to census data
- Limitations of census data
- Other potential things to mention (the affect of aggregation and spatial scale).


### What

Now that we explored the "Who?" question about our data, let's move on to answering the "What?" question. You might have guessed that `sentence_length` and `mandatory_min` are the variables that fall under this category. 

#### How does sentence length correlate with criminal history? {.tabset}

> How do sentence lengths relate to policy? <!--- insert blurb here? -->

To better understand `sentence_length`, we’ll display a histogram for this quantitative variable. A histogram gives us a visual representation of the frequency of values. With R, we can change the width of each bin or choose a number of bins, and then the plot shows us how many sentences fell within each bin range.

##### Histogram

```{r, fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_histogram(aes(x = sentence_length), binwidth = 12) +
  labs(x = "Sentence length in months", y = "Number of individuals")  +
  geom_vline(aes(xintercept = 12), color = "red", linetype = 2) +
  geom_vline(aes(xintercept = 60), color = "red", linetype = 2) +
  geom_vline(aes(xintercept = 120), color = "red", linetype = 2) +
  geom_vline(aes(xintercept = 240), color = "red", linetype = 2)
```

##### Histogram

```{r, fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_histogram(aes(x = sentence_length/12), binwidth = 1) +
  labs(x = "Sentence length in years", y = "Number of individuals") +
  geom_vline(aes(xintercept = 1), color = "red", linetype = 2) +
  geom_vline(aes(xintercept = 5), color = "red", linetype = 2) +
  geom_vline(aes(xintercept = 10), color = "red", linetype = 2) +
  geom_vline(aes(xintercept = 20), color = "red", linetype = 2) +
  labs(title = "Distribution of number of individuals by ")



```

We observe a high frequency over 0, meaning that many individuals who are convicted may in fact receive a sentence of 0 months. At the other extreme, we see cases corresponding to 470 months, representing individuals who either have been given a long sentence, meaning 39 years or possibly a life sentence.

[could further discuss skew, peaks and relate to sentencing table, etc.]

##### Violin Plot

Violin plots are another way of showing the distribution data.

```{r, fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_violin(aes(x = as.factor(criminal_history), 
                  y = sentence_length, 
                  fill = as.factor(criminal_history))) +
  labs(fill = "Criminal History", 
       y = "Sentence length in months", 
       x = "Criminal History")

```

##### Ridge plot

```{r, fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_density_ridges(aes(y = as.factor(criminal_history),
                          x = sentence_length,
                          fill = as.factor(criminal_history))) +
  labs(title = "Distribution of sentence length in months by criminal history",
       subtitle = "Black dashed lines indicate 1, 5, 10, 15, and 20 year sentences",
       fill = "Criminal History",
       x = "Sentence length in months",
       y = "Criminal History") +
  geom_vline(aes(xintercept = 12), color = "black", linetype = 2) +
  geom_vline(aes(xintercept = 60), color = "black", linetype = 2) +
  geom_vline(aes(xintercept = 120), color = "black", linetype = 2) +
  geom_vline(aes(xintercept = 180), color = "black", linetype = 2) +
  geom_vline(aes(xintercept = 240), color = "black", linetype = 2) +
  scale_fill_viridis_d()

```

#### What is the relationship between `all_adjustments` and `sentence_length`?

```{r, fig.width = 7, fig.height = 6}
ggplot(us_sent) +
  geom_point(aes(y = sentence_length,
                 x = all_adjustments, color = as.factor(criminal_history)), alpha = 0.01) +
  labs(x = "All Adjustments",
       y = "Sentence length in months",
       color = "Criminal History") +
  facet_wrap(~ race) +
  scale_color_viridis_d()
```

```{r, fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_point(aes(y = sentence_length,
                 x = all_adjustments, color = as.factor(criminal_history)), alpha = 0.01) +
  labs(x = "All Adjustments",
       y = "Sentence length in months",
       color = "Criminal History") +
  scale_color_viridis_d()
```

#### How does the age range vary with criminal history?

```{r, , fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_density_ridges(aes(y = as.factor(criminal_history),
                          x = age,
                          fill = as.factor(criminal_history))) +
  labs(fill = "Criminal History",
       x = "Age of individual in years",
       y = "Criminal History")
```

### When

Let's explore the time data found in our US sentencing data set. How many columns of data did we have again? This seems like a good place to start digging into what kind of time data we have in this data set. Which tool can we use to find out?


Do we remember?


We do remember: we can use the `names()` routine to remind ourselves.


```{r check the columns, i.e. names}
names(us_sent)
```

We can see the fifteen column names in that output. Awesome. It looks like our time data, as far as "when" these convicted individuals in our data were convicted, is in the `year` column. Let's see what we're working with here by using `distinct()` again.

```{r Segregate and list the data in the column year}
us_sent %>% 
  distinct(year)
```

Excellent. So we know we have years from 2006 to 2020 in our data, or 15 rows worth of data.  It is worth noticing that although 2020-2006 = 14, when we count each year as a whole year of data *inclusively*, we have 15 distinct instances, which the distinct command helpfully listed out for us.

Now, each of these data points is a numeric variable, such as '2010', or '2015'. One way we could check is to use the `sapply()` routine again.

```{r}
sapply(us_sent, is.numeric)
```

In the output we can indeed see `TRUE` for the column `year`. As we progress in our educational journey, we will learn about different numeric variables, including double precision floating point arithmetic numeric variables. Is our `year` data in this format? Let's use the same command with a different argument:


```{r use sapply() to verify the year data is indeed numero}
sapply(us_sent, is.double)
```

Nice, we can see that our `year` data is indeed a double precision floating point number.

Let's do a graph of our `year` data, where we have each year on the vertical y axis and the sentences on the horizontal x axis.  

```{r Bar chart of year data, fig.width = 6, fig.height = 5}
ggplot(us_sent) +
  geom_bar(aes(y = year)) +
  labs(title = "Observations by Year",
       x = "Count",
       y = "Year")
```


If we wanted to use a scatter plot, which could be a bit 'messier', one way to do so would be to use `geom_point()` again:

```{r graph our year data- scatter plot, fig.width = 6, fig.height = 5}
ggplot(us_sent, aes(x = sentence_length, 
                    y = year)) + 
  geom_point() +
  labs(title = "Observations by Year",
       x = "Count",
       y = "Year")
```

### Where

Now that we've explored the "who?", "what?", and "when?" questions about our data set, let's finish off our EDA by understanding the "where?" question. Let's check what districts we have in the data. We can do this using `distinct()` again.

```{r}
us_sent %>%
  distinct(district)
```

You'll notice that certain states are a single district on their own. Other larger states are split into several districts (e.g. New York East, New York North).  

If you live in the United States, do you know which district you live in?

If not, you can look it up at [the United States Department of Justice's website](https://www.justice.gov/ust/locate-your-judicial-district).

Apart from using `distinct` to figure out what districts we have in our data set,  we are also interested in finding out how many sentences were made in each district. We also will want to think about this question in relation to the population of those districts, which can be quite nuanced and we will come back to this later.

Our goals in this next section are to think about ways we can explore the question "Where did those sentences occur?" in a visual way.

**Learning aims**

- introduce factors as a data type
- explain why it might be better to put our names on the y axis instead of the x-axis for many categories and those with longer names. This makes it more readable.
- explain that automatically categorical data will be placed in alphabetical order.
- we can reconfigure the graph using functions from the `forcats` package.
- This is a good example for when you might want to order it in terms of frequency (i.e. the count) using `fct_infreq`
- We may also want to reverse the order to see the districts which have the most sentences at the top of our plot using `fct_rev`.

#### Number of individuals sentences across districts {.tabset}


##### Base Bar Plot

```{r}
ggplot(us_sent) +
  geom_bar(aes(x = district))
```

##### District on the y-axis

```{r}
ggplot(us_sent) +
  geom_bar(aes(y = district))
```

##### Ordered by number of sentences

```{r}
ggplot(us_sent) +
  geom_bar(aes(y = fct_infreq(district)))
```

##### Ordering from high to low

```{r}
ggplot(us_sent) +
  geom_bar(aes(y = fct_rev(fct_infreq(district))))
```

##### Add title and axes labels

```{r}
ggplot(us_sent) +
  geom_bar(aes(y = fct_rev(fct_infreq(district)))) +
  labs(title = "Number of individuals sentenced at the federal district court level for each district from x to x",
       y = "Federal District Court",
       x = "Number of individuals")
```

##### Making the plot more readable with subsetting

```{r}
ggplot(us_sent) +
  geom_bar(aes(y = fct_rev(fct_infreq(district)))) +
  labs(title = "Number of individuals sentenced at the federal district court level for each district from x to x",
       y = "Federal District Court",
       x = "Number of individuals")
```
```{r}
ggplot(us_sent) +
  geom_bar(aes(y = fct_infreq(district))) +
  labs(title = "Number of individuals sentenced at the federal district court level for each district from x to x",
       y = "Federal District Court",
       x = "Number of individuals")
```

##### Exploring the census data

**To Do**:

- If we want to get out the district populations, we will probably need to download the data by the county level.
- Decision points: we could download the data in categories by age and sex and race: e.g. 5-9, 10-14 etc. This might get a little

```{r census-data-import, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}

# apply unique census api key
census_api_key("5177724b01a7fe4714097e711cb95230c37cfce7", overwrite = TRUE)

# import census data
## guide to spatial units: https://api.census.gov/data/2016/acs/acs5/geography.html
## variable of interest --> population
## vars <- load_variables(year = 2013,
                      # data set = "acs5",
                      # cache = TRUE)

# B02001_001: Total
# B03002_003: White alone (Not Hispanic or Latino)
# B03002_004 Black or African American alone (Not Hispanic or Latino)
# B03002_012: Hispanic or Latino
# B03002_005: Native American alone (Not Hispanic or Latino)
# B03002_006: Asian alone (Not Hispanic or Latino)
# B03002_007: Native Hawaiian or Pacific Islander alone (Not Hispanic or Latino)
# B03002_009: Multiple Races (Not Hispanic or Latino)
# B03002_008: Other (Not Hispanic or Latino)

#census_place_df <- get_acs(geography = "state", variables = c("B01003_001E"), geometry = TRUE, year = 2010)

```
