---
title: "Sentencing Tutorial"
author: "Laurie"
date: "2023-07-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
```

First you will want to read in your data. You can do this using `read_csv`. To open the data you can "uncomment" (delete the hash tag #) to view the data file. Note that you will need to "recomment" this out in order to knit your file again.

```{r read in data, message=FALSE, warning=FALSE, echo=FALSE}

us_sent <- read_csv("data/cleaned_data_renamed.csv")
#View(us_sent)
```

Let's inspect our variables which are in the columns.

```{r check column names}
names(us_sent)

glimpse(us_sent)
```
Let's check what districts we have in the data.

```{r}
us_sent %>%
  distinct(district)
```
You'll notice that certain states are a single district on their own. Other larger states are split into several districts (e.g. New York East, New York North). If you live in the United States. 

Q: Can you identify which district you live in?

### What does each row represent?

In a tidy dataset, each row is an observation, each column is a variable, and each cell is a value.

In this case each row represents a single court sentence at the federal court level


### 


###


###

